---
title: <span style="font-family:Georgia;">Differential expression analysis in pre-treatment samples</span>
author: "Víctor Jiménez Martínez"
date: "2022"
output: 
  html_document:
    toc: true
    theme: united
---

## Introduction

The main goal of this differential expression analysis is to compare RNA-seq expression data between responder and non-responder patients and explore the differences in their gene expression. The analyses are performed with protein-coding genes. 

## Libraries

Libraries needed for the analyses are loaded below.

```{r, results="hide", message=FALSE, warning=FALSE}
library(edgeR)  # to build MA plots
library(pheatmap)  # to build a heatmap
library(stats)
library(ggplot2)  # for visualization
library(EDASeq)  # for RLE plots
library(limma)  # for differential expression analysis
library(DESeq2) # for differential expression analysis
library(sva)  # for surrogate variable analysis
library(EnhancedVolcano)  # for Volcano plot
library(clusterProfiler)  # for enrichment analysis
library(org.Hs.eg.db)  # Homo sapiens annotation
library(OmicCircos)  # for circosPlot
library(GWASTools) # for qqplot
library(fgsea) # GSEA
library(tibble) # to manage data
library(simplifyEnrichment)
```

## Loading data

First of all, data is loaded and set for the analyses.

```{r}
# MCPCounter results
mcp <- read.delim("/omics/groups/OE0436/data/jimenez/outputs/rnaseq/deconvolution/mcp_decon_immune_pre.tsv")

# Raw counts
counts <- read.delim("/omics/groups/OE0436/data/jimenez/data/rnaseq/rnaseq_counts_met1_protein_coding.tsv")
colnames(counts)[1:5] <- sapply(colnames(counts)[1:5], function(x) gsub("^X", "", x)) # adapt three first PIDs

# Response to treatment information
pheno <- read.csv(file = "/omics/groups/OE0436/data/jimenez/data/PID_treatment.csv", 
                  header = TRUE, sep = ";", row.names = "PID")
pheno <- pheno[colnames(counts), ] # order the same and filter for those samples we have in rnaseq
```

Set SD and PD as non-responders and PR as responders.

```{r}
colnames(pheno) = c("therapy", "evolution") # change names

pheno <- pheno %>% 
  mutate(response = ifelse(evolution == "PR", "responders", "non-responders"))
```

Finally, counts matrices are filtered to remove background transcriptomic noise and genes that are not expressed. First of all, all rows sums must be higher than 10 in order to ensure there is a minimal amount of counts per gene, and second of all, at least 80% of samples per condition must be different to 0 counts. 

```{r}
# Filter counts table to remove some background "noise"
# Row sums must be higher than 10 in both groups in order to ensure there are enough counts
# counts <- counts[rowSums(counts) >10, ]
counts <- counts[rowSums(counts[,rownames(pheno)[pheno$response == "responders"]]) > 10 &
                    rowSums(counts[,rownames(pheno)[pheno$response == "non-responders"]]) > 10, ]

cat(dim(counts)[1], 
    "genes with more than 10 counts in both groups")

# Keep genes that have counts in at least 80% of each group
# Compute number of samples per experimental group and minimum percent
n.samplespergroup <- table(pheno$response) 
percent <- 0.8 # define the 80%
samples.min <- round(n.samplespergroup*(1-percent), 0) # number of samples that represents the 20%, the maximum amount of 0 allowed

# Data frame with the number of 0 in each gene and sample
df_0 <- data.frame(row.names = rownames(counts), 
                   "non-responders" = rep(0, length(rownames(counts))), 
                   "responders" = rep(0, length(rownames(counts))))
for (i in 1:2){
  c <- counts[ , which(colnames(counts) %in% rownames(pheno[pheno$response == unique(pheno$response)[i], ]))] # subset counts per group
  how_many_0 <- apply(c, 1, function(x) sum(x == 0)) # compute how many 0 per gene
  df_0[, i] <- as.vector(how_many_0)
}

# Get genes that have counts in at least 80% of the samples of at least one group
resta <- apply(df_0, 1,  function(x) {any(x < as.vector(samples.min))}) # substract the 80% computed for each group to the counts found for each gene, if it is > 0 it means that at least 60% of samples counted that gene
idx_80 <- rownames(counts)[which(resta == TRUE)]

cat(length(idx_80), "out of", dim(counts)[1], 
    "genes counted in 80% of the samples of at least one group")

counts <- counts[idx_80, ]

# Before applying DEseq2, normalized data will be checked to ensure normalization is accomplished in our set and continue with the analysis
# TPM counts are loaded and the selected genes in counts are selected
tpm <- read.delim("/omics/groups/OE0436/data/jimenez/data/rnaseq/rnaseq_tpm_met1_protein_coding.tsv")
tpm <- tpm[rownames(counts), ]
colnames(tpm)[1:5] <- sapply(colnames(tpm)[1:5], function(x) gsub("^X", "", x)) # adapt three first PIDs
```

When comparing responders vs. non-responders, estimation of immune cell population abundances through MCPCounter tool will be included as covariates. But, first of all, abundances will be normalized.

```{r}
# Z-score normalization
zmcp <- (mcp[,-29] - apply(mcp[,-29], 1, mean)) / apply(mcp[,-29], 1, sd)

# Check normalization with RLE
plotRLE(as.matrix(t(mcp[,-29])), outline = FALSE, ylim = c(-2, 2), main = "pre-treatment MCPCounter raw estimated relative abundance")
plotRLE(as.matrix(t(zmcp)), outline = FALSE, ylim = c(-2, 2), main = "post-treatment MCPCounter z-score estimated relative abundance")
plotRLE(as.matrix(log2(mcp[,-29])), outline = FALSE, ylim = c(-2, 2), main = "Log2")

# While log2 normalization seems to have less dispersion, zcount scores reduce sd and mean in the variables, which makes it a better option for fiting the variables into GLM regressions in DESeq2

# Including the es of populations in pheno for further analysis
pheno <- cbind(pheno, zmcp)

# Fixing therapy information
pheno$therapy[pheno$therapy == "Pembro + T-VEC"] = "Pembro + TVEC"
pheno$therapy[pheno$therapy == "Ipi/Nivo "] = "Ipi/Nivo"
pheno$therapy[pheno$therapy == "Pembro + TVEC"] = "Pembro_and_TVEC"
pheno$therapy[pheno$therapy == "Ipi/Nivo"] = "Ipi_Nivo"

# Adding batch information
pheno$batch <- as.factor(c(2, 1, 1, 1, 2, 2, 2, 2, 1, 2, 1, 1, 2, 1, 1, 1, 1, 1, 1, 2, 2))
```

## Quality control and normalization

To perform a differential expression analysis, we must check normalized data and its variation. Even though DEseq2 requires raw counts, it performs a normalization during its process. 

Distribution of normalized counts can be checked. Assuming that, between two individuals, most genes are not differentially expressed, the difference of log-ratios between individuals should be centered at 0 when correctly normalized

```{r}
MbyTPM <- log2(tpm[, 1] - tpm[, 3])

hist(as.numeric(MbyTPM), xlab = "log2-ratio", main = "pre-treatment TPM differences between 2 samples")
abline(v = 0, col = "red")
```

To further check if data is correctly normalized a MA-plot. The plot visualizes the differences between measurements taken in two samples, by transforming the data onto M (log ratio) and A (mean average) scales, then plotting these values. That is, it represents the difference of count log-rations between individuals against the mean of those log-ratios.The red line would show the expected values.

```{r}
maPlot(tpm[, 1], tpm[, 2], 
       pch = 19, cex = 0.5, ylim = c(-8, 8), 
       allCol = "darkgray", lowess = TRUE, 
       xlab = expression(A == log[2](sqrt(Sample1 %.% Sample2))), 
       ylab = expression(M == log[2](Sample1/Sample2)))
grid(col = "black")
title("MA plot of TPM differences between 2 samples")
```

Finally, dispertion can be checked with RLE (Relative Log Expression) plots. RLE plots are a tool for visualizing unwanted variation in high dimensional data, showing the dispersion of the samples.

```{r}
par(mfrow = c(2, 1))

plotRLE(as.matrix(counts), outline = FALSE, ylim = c(-2, 2), main = "RLE of pre-treatment raw counts")

plotRLE(as.matrix(tpm), outline = FALSE, ylim = c(-2, 2), main = "RLE of pre-treatment TPM")
```

## Exploratory analysis

A heatmap and a PCA are performed to check how the samples cluster according to their response.

### Clustering through heatmap

For clustering of data, a heatmap is built. That way, similarity between individuals according to quantified expression is addressed. To save resources and time, only most variable genes are used for the heatmap.

```{r}
# Selection of top 100 most variable genes Compute the variance of each gene across
# samples
V <- apply(tpm, 1, var)

# Results are sorted by variance in decreasing order and top 100 genes are selected
selectedGenes <- names(V[order(V, decreasing = T)][1:100])

# Heatmap
pheatmap(tpm[selectedGenes, ], scale = "row", annotation_col = pheno[,c(2,3,32)], 
         show_rownames = FALSE, show_colnames = FALSE, 
         annotation_legend = T, annotation_names_col = F)
```

They do not seem to cluster

### PCA
A PCA is performed to see the distribution of individuals according to response to treatment and check again if any clustering can be seen.

```{r}
# Compute PCA
pca_data <- prcomp(t(tpm), retx=TRUE, center=TRUE, scale=FALSE)
summary(pca_data)$importance[,c(1,2)]

# Scores data
scores <- as.data.frame(pca_data$x[, c("PC1", "PC2")]) # loading coeficients
scores$Response <- pheno$response # response to treatment 
scores$batch <- pheno$batch

# Plot
pdf("/omics/groups/OE0436/data/jimenez/outputs/rnaseq/DEA/protein_coding_genes/PCA_expression_tpm_protein_coding.pdf")

scores.plot <- ggplot(data = scores, 
                      aes(x = PC1, 
                          y = PC2, 
                          colour = Response, 
                          label = rownames(scores))) + 
  geom_point(alpha = I(0.7), size = 4) + 
  geom_hline(yintercept = 0)+
  geom_vline(xintercept = 0)+
  xlab(paste0("PC1 (", round(summary(pca_data)$importance[2,1], 2) * 100, "%)"))+
  ylab(paste0("PC2 (", round(summary(pca_data)$importance[2,2], 2) * 100, "%)"))+
  stat_ellipse() + scale_color_manual(values = c("red", "green")) +
  theme_bw() + ggtitle("PCA with expression data (TPM) in pre-treatment samples")

scores.plot

dev.off()
```

They do not cluster.

## Accounting for unwanted variability 

Once data is loaded and checked, Differential Expression Analysis may be performed. However, it is important to take into account unwanted variability that may biass the results and inflate the p-values. P-values are expected to follow a uniform distribution. Extra variability may inflate p-values and give more significanies than we should have (false positives). Thus, a first iteration of the DEseq2^1^ is performed to check on p-value distribution and if there needs to be correction due to inflation.

The covariates that will be used for the analyses will include the therapies and the estimated abundance of a set of immune cells. These populations are selected from the results of MCPCounter deconvolution and are the ones that showed significant differences in post-treatment samples, considering they change during the treatment. These populations are: Activated CD4 T cell, Activated CD8 T cell, Central Memory CD8 T cell, Effector Memory CD8 T cell, Macrophage, Immature B cell, Mast cell, Natural Killer cell, Regulatory T cell, T follicular helper cell, Type 1 T helper cell. According to the results commented in *deconvolotuion.Rmd*, it would seem that CD8+ populations could constitute a potential biomarker for response to immune checkpoint treatment, something partially supported by the change during treatment. In that sense, the abundance of these populations will be included as covariates for the differential expression analysis. 


```{r}
# Create a DESeq dataset object from the count matrix and the colData
# All the variables in pheno are included
designFormula <- as.formula(~ therapy + response + batch + Activated.CD8.T.cell + 
                              Central.memory.CD8.T.cell  + Effector.memeory.CD8.T.cell) 
                            
dds <- DESeqDataSetFromMatrix(countData = round(counts),
                              colData = pheno, 
                              design = designFormula)

# We can use the DESeq () function of DESeq2 , which is a wrapper function that
# implements estimation of size factors to normalize the counts, estimation of dispersion values,
# and computing a GLM model based on the experimental design formula
dds <- DESeq(dds)
# dds <- nbinomWaldTest(dds, maxit = 300) 
# 13 genes do not converge, might be because they bear low counts and the filter did not get rid off them

# Compare and contrast the samples based on different variables of interest
DEresults <- results(dds, contrast = c("response", 'responders', 'non-responders'))
DEresults <- DEresults[order(DEresults$pvalue),]

# p-value distribution
ggplot(data = as.data.frame(DEresults), aes(x = pvalue)) +
  geom_histogram(bins = 100) + theme_bw()

# qq-plots
par(mfrow = c(1, 2))
GWASTools::qqPlot(DEresults$pvalue, main = "qq-plot with p-values")
GWASTools::qqPlot(DEresults$padj, main = "qq-plot with adjusted p-values")
```

### SVA
The qq-plot shows the expected distribution of p-values vs. the actual distribution obtained. We can see our p-values are a little bit inflated. Thus, a SVA analysis to catch extra variability to adjust to is performed. 

Surrogate Variable Analysis (SVA)^2^ was performed with sva R package to account for unwanted existing variability (except the one driven by our condition) and adjust for it in the DESeq2 model. By defining two models, the method esti-mates artificial surrogate variables from raw counts that were then included in the DESeq2 design matrix. 

```{r}
mod1 <- model.matrix( ~ therapy + response + batch + Activated.CD8.T.cell + 
                              Central.memory.CD8.T.cell  + Effector.memeory.CD8.T.cell, 
                      data = pheno)

                  
mod0 <- model.matrix( ~ therapy + batch + Activated.CD8.T.cell + 
                              Central.memory.CD8.T.cell  + Effector.memeory.CD8.T.cell, 
                      data = pheno) # model with all the covariates but the one of interest (response)


# Then, the surrogate variables are estimated with
sv <- svaseq(as.matrix(counts), mod1, mod0) # does not work without filtering genes...

# We include these surrogate variables in the pheno data frame
for (i in 1:4){
  pheno[,32+i] <- sv$sv[,i]
}
colnames(pheno)[33:36] <- c("sv1", "sv2", "sv3", "sv4")
```

Let's try again.

```{r}
# Create a DESeq dataset object from the count matrix and the colData
# All the variables in pheno are included
designFormula <- as.formula(~ therapy + response + batch + Activated.CD8.T.cell + 
                              Central.memory.CD8.T.cell  + Effector.memeory.CD8.T.cell +
                              sv1 + sv2 + sv3 + sv4) 
                            
dds_sv <- DESeqDataSetFromMatrix(countData = round(counts),
                              colData = pheno, 
                              design = designFormula)

# We can use the DESeq () function of DESeq2 , which is a wrapper function that
# implements estimation of size factors to normalize the counts, estimation of dispersion values,
# and computing a GLM model based on the experimental design formula
dds_sv <- DESeq(dds_sv)

# dds_sv <- nbinomWaldTest(dds_sv, maxit = 741) # to converge all genes I increase the number of iterations

# 103 genes do not converge

# # To converge all iterations
# dds_sv <- estimateSizeFactors(dds_sv)
# dds_sv <- estimateDispersions(dds_sv)
# dds_sv <- nbinomWaldTest(dds_sv, maxit = 100)

# Compare and contrast the samples based on different variables of interest
DEresults_sv <- results(dds_sv, contrast = c("response", 'responders', 'non-responders'))
DEresults_sv <- DEresults_sv[order(DEresults_sv$pvalue),]

# p-value distribution
ggplot(data = as.data.frame(DEresults_sv), aes(x = pvalue)) +
  geom_histogram(bins = 100) + theme_bw()

# qq-plots
par(mfrow = c(1, 2))
pdf("/omics/groups/OE0436/data/jimenez/outputs/rnaseq/DEA/protein_coding_genes/p-value_distribution_DEA_SV_pre.pdf")

GWASTools::qqPlot(DEresults_sv$pvalue, main = "qq-plot with p-values after SVA")

dev.off()

GWASTools::qqPlot(DEresults_sv$padj, main = "qq-plot with adjusted p-values")

pdf("/omics/groups/OE0436/data/jimenez/outputs/rnaseq/DEA/protein_coding_genes/p-value_distribution_DEA_SV_pre.pdf")

ggplot(data = as.data.frame(DEresults_sv), aes(x = pvalue)) +
  geom_histogram(bins = 100)
GWASTools::qqPlot(DEresults_sv$pvalue, main = "qq-plot with p-values")

dev.off()
```

### ComBat

Since SVA did not work alone as expected adn p-values are still inflated, ComBat can be applied. However, it is a very aggressive method that transforms data and may be not the best option in our case, where inflation is not extremely high.

```{r}
# Apply ComBat for RNA-seq data
combat_counts <- ComBat_seq(as.matrix(counts), batch = pheno$batch)
```

## Differential expression analysis

With the modified ComBat counts and estimated Surrogate Variables accounting for extra and unwanted variability, differences in expresison may be finally assessed properly.

```{r}
# # Create a DESeq dataset object from the count matrix and the colData
# # All the variables in pheno are included
# designFormula <- as.formula(~ therapy + response + batch + Activated.CD8.T.cell + 
#                               Central.memory.CD8.T.cell  + Effector.memeory.CD8.T.cell + 
#                               sv1 +sv2 + sv3 + sv4) # still including sv 
#                             
# dds_combat <- DESeqDataSetFromMatrix(countData = round(combat_counts),
#                               colData = pheno, 
#                               design = designFormula)
# 
# # We can use the DESeq () function of DESeq2 , which is a wrapper function that
# # implements estimation of size factors to normalize the counts, estimation of dispersion values,
# # and computing a GLM model based on the experimental design formula
# dds_combat <- DESeq(dds_combat)
# 
# # 105 rows do not converge again
# 
# # dds_combat <- nbinomWaldTest(dds_combat, maxit = 298.5) # to converge all genes I increase the number of iterations 
# 
# # Compare and contrast the samples based on different variables of interest
# DEresults_combat <- results(dds_combat, contrast = c("response", 'responders', 'non-responders'))
# DEresults_combat <- DEresults_combat[order(DEresults_combat$pvalue),]
# 
# # p-value distribution
# ggplot(data = as.data.frame(DEresults_combat), aes(x = pvalue)) +
#   geom_histogram(bins = 100)
# 
# # qq-plots
# par(mfrow = c(1, 2))
# GWASTools::qqPlot(DEresults_combat$pvalue, main = "qq-plot with p-values")
# GWASTools::qqPlot(DEresults_combat$padj, main = "qq-plot with adjusted p-values")
```

After all, methods to account for extra variability are not cahnging very much the original results in the distribution of the p-values. Thus, I discard ComBat results, since they do a very aggressive modification of the data. Between adding SVA variables or not, the difference is the collpase of some p-values to 0, but the distribution is not greatly changed. To catch those values, then, I use SVA results. 

```{r}
# Exploring the top DEG
# Top differentially expressed genes
DEresults_sv <- DEresults_sv[rowSums(is.na(DEresults_sv)) == 0, ] # remove NA rows
top <- DEresults_sv[DEresults_sv$padj < 0.05, ] # 153 differentially expressed genes
# DEresults <- DEresults[rowSums(is.na(DEresults)) == 0, ] # remove NA rows
# top <- DEresults[DEresults$padj < 0.05, ] # 117 differentially expressed genes

# Save table
write.table(top, file = "/omics/groups/OE0436/data/jimenez/outputs/rnaseq/DEA/protein_coding_genes/DEG_protein_coding_pre_nofilter.tsv", sep = "\t", col.names = TRUE, row.names = TRUE)


# Normalized counts from differentially expressed genes
dif <- tpm[rownames(top), ]

# Heatmap
pdf("/omics/groups/OE0436/data/jimenez/outputs/rnaseq/DEA/protein_coding_genes/heatmap_DEG_protein_coding_pre.pdf")

pheatmap(as.matrix(dif[,order(pheno$response)]), scale = "row", 
         annotation_col = data.frame(row.names = rownames(pheno)[order(pheno$response)], 
                                     "Response" = pheno$response[order(pheno$response)]),
         annotation_colors = list(Response= c(`non-responders` = "red", responders= "green")),
         show_rownames = FALSE, show_colnames = TRUE, # legend_labels = c("Scaled expression"),
         cluster_cols = FALSE, annotation_legend = T, annotation_names_col = F,
         name = "Scaled Expression (TPM)")

dev.off() 

# Volcano plot
pdf("/omics/groups/OE0436/data/jimenez/outputs/rnaseq/DEA/protein_coding_genes/volcano_plot_protein_coding_pre.pdf")

EnhancedVolcano(DEresults_sv, lab = rownames(DEresults_sv), x = "log2FoldChange", y = "padj",
                pCutoff = 0.05, FCcutoff = 1, 
                # selectLab =  rownames(comparison),
                title = "Volcano plot", 
                subtitle = "pre-treatment responders vs nonresponders differential expression analysis",
                legendLabSize = 10, legendIconSize = 3, legendPosition = "right", labSize = 3)
                #legendLabels=c('Not sig.','Log (base 2) FC','p-value', 'p-value & Log (base 2) FC'))

dev.off()

# PCA
# Compute PCA
pca_data <- prcomp(t(dif), retx=TRUE, center=TRUE, scale=FALSE)
summary(pca_data)$importance[,c(1,2)]

# Scores data
scores <- as.data.frame(pca_data$x[, c("PC1", "PC2")]) # loading coeficients
scores$Response <- pheno$response # response to treatment 
scores$batch <- pheno$batch

# Plot
pdf("/omics/groups/OE0436/data/jimenez/outputs/rnaseq/DEA/protein_coding_genes/PCA_deg_pre_protein_coding.pdf")

scores.plot <- ggplot(data = scores, 
                      aes(x = PC1, 
                          y = PC2, 
                          colour = Response, 
                          label = rownames(scores))) + 
  geom_point(alpha = I(0.7), size = 4) + 
  geom_hline(yintercept = 0)+
  geom_vline(xintercept = 0)+
  xlab(paste0("PC1 (", round(summary(pca_data)$importance[2,1], 2) * 100, "%)"))+
  ylab(paste0("PC2 (", round(summary(pca_data)$importance[2,2], 2) * 100, "%)"))+
  stat_ellipse() + scale_color_manual(values = c("red", "green")) +
  theme_bw() + ggtitle("PCA with pre-treatment differentially expressed genes expression (TPM)")

scores.plot

dev.off()

```

With the results, we can proceed with GSEA and enrichment analyses. 

## Functional analyses

Enrichment analysis is performed to identify over-represented gene ontologies (GO^3^ Database) and pathways (KEGG^4^ database) in responders and non-responders from their differential expressed genes (adjusted p-value < 0.05). For that purpose, clusterProfiler^5^ R package was implemented, which conducts enrichment tests based on hypergeometric distribution and adjusts p-values and computes q-values. 

### Enrichment

```{r}
# Subset for upregulation according to condition (the contrast was responders vs nonresponders)
responders_up <- top[top$log2FoldChange > 0, ]
nonresponders_up <- top[top$log2FoldChange < 0, ]

# Obtain genes 
deGenes <- rownames(top)
deGenes_responders <- rownames(responders_up)
deGenes_non <- rownames(nonresponders_up)

# Obtain EntrezID of deg
deGenes <- bitr(deGenes, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db")
deGenes_responders <- bitr(deGenes_responders, fromType="SYMBOL", 
                           toType="ENTREZID", OrgDb="org.Hs.eg.db")
deGenes_non <- bitr(deGenes_non, fromType="SYMBOL", 
                    toType="ENTREZID", OrgDb="org.Hs.eg.db")

# Given our list, the function searches for human Biological Processes associated in GO and KEGG
# with an adjusted p-value cut-off 
# ego_top <- enrichGO(gene = deGenes$ENTREZID, 
#                     ont = "BP", 
#                     keyType = "ENTREZID",
#                     OrgDb = "org.Hs.eg.db", 
#                     readable = TRUE, 
#                     pvalueCutoff = 0.05)
# 
# ek_top <- enrichKEGG(gene = deGenes$ENTREZID,
#                      pvalueCutoff = 0.5,
#                      qvalueCutoff = 0.5,
#                      organism = "hsa")
# 
# enrichplot::dotplot(ego_top, title = "GO responders vs non-responders differentially expressed genes", showCategory = 10)
# enrichplot::dotplot(ek_top, title = "KEGG responders vs non-responders differentially expressed genes", showCategory = 10)

ego_responders <- enrichGO(gene = deGenes_responders$ENTREZID, 
                   ont = "BP", 
                   keyType = "ENTREZID",
                   OrgDb = "org.Hs.eg.db", 
                   readable = TRUE, 
                   pvalueCutoff = 0.2)

ek_responders <- enrichKEGG(gene = deGenes_responders$ENTREZID,
                    pvalueCutoff = 0.5,
                    qvalueCutoff = 0.5,
                    organism = "hsa")

ego_non <- enrichGO(gene = deGenes_non$ENTREZID, 
                   ont = "BP", 
                   keyType = "ENTREZID",
                   OrgDb = "org.Hs.eg.db", 
                   readable = TRUE, 
                   pvalueCutoff = 0.2)

ek_non <- enrichKEGG(gene = deGenes_non$ENTREZID,
                    pvalueCutoff = 0.5,
                    qvalueCutoff = 0.5,
                    organism = "hsa")

# Table with all of them
ego_responders@result$Response <- "responders"
ego_responders@result$Type <- "GO term" 

ego_non@result$Response <- "non-responders"
ego_non@result$Type <- "GO term" 

ek_responders@result$Response <- "responders"
ek_responders@result$Type <- "KEGG pathway" 

ek_non@result$Response <- "non-responders"
ek_non@result$Type <- "KEGG pathway" 

enrichment <- rbind(ego_responders@result, ego_non@result, ek_responders@result, ek_non@result)

write.table(enrichment, 
            file = "/omics/groups/OE0436/data/jimenez/outputs/rnaseq/DEA/protein_coding_genes/enrichment/deg_enrichment.csv", quote = FALSE, sep = ",", row.names = TRUE, col.names = TRUE)


# Some exploration
ego_responders@result[ego_responders@result$p.adjust < 0.05, ] # 67 terms significantly associated
terms_resp <- c("positive regulation of inflammatory response",
                "neutrophil migration", "granulocyte migration", "leukocyte migration",
                "neutrophil chemotaxis", "leukocyte chemotaxis", "granulocyte chemotaxis",
                "leukocyte aggregation", "positive regulation of defense response", 
                "leukocyte cell-cell adhesion", "cytokine-mediated signaling pathway",
                "regulation of inflammatory response", "myeloid leukocyte migration",
                "positive regulation of interleukin-8 production",
                "leukocyte migration involved in inflammatory response",
                "positive regulation of T cell activation")
ego_resp_immune <- subset_enrichResult(ego_responders, which(ego_responders@result$Description %in% terms_resp)) # subset

ek_responders@result[ek_responders@result$p.adjust < 0.05, ] # Only cytokine-cytokine receptor interaction is significantly upregulated (hsa04060)

ego_non@result[ego_non@result$p.adjust < 0.1, ] # no terms with p-adjusted < 0.05 and no interesting terms...

ek_non@result[ek_non@result$p.adjust < 0.1, ] # no terms associated...

pdf("/omics/groups/OE0436/data/jimenez/outputs/rnaseq/DEA/protein_coding_genes/enrichment/dotplots_enrichment_pre_protein_coding_genes.pdf")

enrichplot::dotplot(ego_responders, title = "GO terms enriched in responders significantly upregulated genes", showCategory = 20)
enrichplot::cnetplot(ego_resp_immune)

enrichplot::dotplot(ek_responders, title = "KEGG pathways enriched in responders significantly upregulated genes", showCategory = 10)

enrichplot::dotplot(ego_non, title = "GO terms enriched in non-responders significantly upregulated genes", showCategory = 10)
enrichplot::dotplot(ek_non, title = "KEGG pathways enriched in non-responders significantly upregulated genes", showCategory = 10)

dev.off()

# cytokine-cytokine receptor interaction
pathview(gene.data = deGenes_responders$ENTREZID, pathway.id = "hsa04060", species = "hsa")
pathview(gene.data = deGenes_responders$ENTREZID, pathway.id = "hsa04060", 
                species = "hsa", kegg.native = F)

```

There seems to be an enrichment of immunologic functions in both groups. 

### GSEA

Another way to check biological significance of genes through Gene-set enrichment analysis.

```{r}
data(examplePathways) # gene sets of pathways provided by the package

# Prepare genes
DEresults_sv$SYMBOL <- rownames(DEresults_sv)
DEres2 <- as.data.frame(DEresults_sv) %>% 
  dplyr::select(stat, SYMBOL) %>% 
  na.omit() %>% 
  distinct() %>% 
  group_by(SYMBOL) %>% 
  summarize(stat = mean(stat))

ranks <- deframe(DEres2)
head(ranks, 20)

# Load the pathways into a named list
# Hallmark gene set from MSigDB
pathways.hallmark <- gmtPathways("/omics/groups/OE0436/data/jimenez/data/h.all.v7.5.1.symbols.gmt")

# Immunge gene sets from MSigDB
immune.hallmark <- gmtPathways("/omics/groups/OE0436/data/jimenez/data/c7.immunesigdb.v7.5.1.symbols.gmt")

# Let's subset for uninteresting hallmarks: there are too many, either way...
out <- unique(c(grep("VACCIN*", names(immune.hallmark)), grep("VIR", names(immune.hallmark)),
         grep("CULT", names(immune.hallmark)), grep("DAY", names(immune.hallmark)),
         grep("VITRO", names(immune.hallmark)), grep("KO", names(immune.hallmark)),
         grep("H_", names(immune.hallmark)), grep("WT", names(immune.hallmark)),
         grep("TREAT", names(immune.hallmark))))
immune.hallmark <- immune.hallmark[-out]

# GSEA
fgseaRes <- fgsea(pathways = pathways.hallmark, 
                  stats = ranks, 
                  nperm = 1000)

fgseaRes_immune <- fgsea(pathways = immune.hallmark, 
                  stats = ranks, 
                  nperm = 1000)

# Order results
fgseaResTidy <- fgseaRes %>%
  as_tibble() %>%
  arrange(desc(NES))

fgseaResTidy_immune <- fgseaRes_immune %>%
  as_tibble() %>%
  arrange(desc(NES))

# Show in a nice table:
fgseaResTidy %>% 
  dplyr::select(-leadingEdge, -ES, -nMoreExtreme) %>% 
  arrange(padj) %>% 
  DT::datatable()

fgseaResTidy_immune %>% 
  dplyr::select(-leadingEdge, -ES, -nMoreExtreme) %>% 
  arrange(padj) %>% 
  DT::datatable()

pdf("/omics/groups/OE0436/data/jimenez/outputs/rnaseq/DEA/protein_coding_genes/enrichment/fgsea_protein_coding_genes.pdf")

ggplot(fgseaResTidy, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill = padj < 0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="Hallmark pathways from GSEA") + 
  theme_bw()

ggplot(fgseaResTidy_immune[1:20, ], aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill = pval < 0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="Immune hallmarks NES from GSEA") + 
  theme_minimal()

dev.off()
```

There are enriched immunological pathways and functions.

## Further exploration

According to paper :
"We reasoned that the immunogenicity of the tumor can be represented by the cytolytic activity  (estimated using the expression of granzyme A (GZMA) and perforin (PRF1) according to Rooney et al., 2015), because this is the ultimate effector mechanism in the cancer immunity cycle"

```{r}
DEresults_sv["GZMA", ] # it's not significantly different
DEresults_sv["PRF1", ] # it's not significantly different
DEresults_sv["CD274", ] # PD-L1 gene is not significantly different
DEresults_sv["HLA-A", ] # it's not significantly different
DEresults_sv[grep("HLA", rownames(DEresults_sv)), ] # no
DEresults_sv["TNF", ] # it's not significantly different
DEresults_sv["IFNG", ] # it's not significantly different
DEresults_sv[grep("CXCL", rownames(DEresults_sv)), ] # a bunch of them
DEresults_sv[grep("IL", rownames(DEresults_sv)), ] # IL1RAPL1 is significant
DEresults_sv[grep("SOCS", rownames(DEresults_sv)), ] # none of them are significantly different
DEresults_sv["TCF7", ] # it's not significantly different

# Interesting genes
int_genes <- c("GZMA", "PRF1", "CD274", "TNF", "IFNG", "CTLA4")
tpm_int <- tpm[int_genes, ] %>% mutate(GENE = int_genes)
rownames(tpm_int) <- 1:nrow(tpm_int)

tpm_int <- pivot_longer(tpm_int, cols = colnames(tpm), names_to = "PID", values_to = "Expression") %>%
  mutate(Response = ifelse(PID %in% rownames(pheno)[pheno$response == "responders"], "responders",
                           "non-responders"))


ggplot(tpm_int, 
       aes(x = Response, y = Expression)) + 
  geom_boxplot(aes(fill = Response), alpha = 2) + geom_point() + 
  scale_fill_manual(values = c("red", "green")) +
  theme_bw() + ylab("TPM expression values") + 
  xlab("Expression of known biomarkers for metastasis1 samples (pre-treatment)") +
  facet_wrap(~ GENE, scales = "free_y") +
  theme(axis.text.x = element_blank(), 
        strip.background = element_rect(fill = "white", color = "black"),
        strip.text.x = element_text(size = 12)) + 
  stat_compare_means(method = "t.test", paired = FALSE, vjust = TRUE, size = 3, label.x = 1.25)



# EnhancedVolcano(DEresults_combat, lab = rownames(DEresults_combat), x = "log2FoldChange", y = "padj",
#                 pCutoff = 0.05, FCcutoff = 10, 
#                 selectLab = c("GZMA", "PRF1", "CD274", "IFNG", "TNF", "CXCL3", "IL-17", "TCF7"),
#                 title = "Volcano plot", 
#                 subtitle = "responders vs nonresponders differential expression analysis",
#                 legendLabSize = 10, legendIconSize = 3, legendPosition = "right", labSize = 3,
#                 parseLabels = TRUE)
#                 #legendLabels=c('Not sig.','Log (base 2) FC','p-value', 'p-value & Log (base 2) FC'))

```


## Bibliography
1. Love, M. I., Huber, W. & Anders, S. Moderated estimation of fold change and dispersion for RNA-seq data with DESeq2. Genome Biol 15, 1–21 (2014).
2. Leek, J. T. & Storey, J. D. Capturing Heteroge-neity in Gene Expression Studies by Surrogate Variable Analysis. PLoS Genet 3, e161 (2007).
3. Gene Ontology Consortium, T. et al. Gene On-tology: tool for the unification of biology NIH Public Access Author Manuscript. Nat Genet 25, 25–29 (2000).
4. Kanehisa, M., Goto, S., Furumichi, M., Tanabe, M. & Hirakawa, M. KEGG for representation and analysis of molecular networks involving diseases and drugs. doi:10.1093/nar/gkp896.
5. Yu, G., Wang, L.-G., Han, Y. & He, Q.-Y. clus-terProfiler: an R Package for Comparing Bio-logical Themes Among Gene Clusters. doi:10.1089/omi.2011.0118.















