---
title: <span style="font-family:Georgia;">Analysis of MutScape significant genes</span>
author: "Víctor Jiménez Martínez"
date: "2022"
output: 
  html_document:
    toc: true
    theme: united
---

## Introduction

The main goal of this script is to analyze and represent results from significantly mutated genes obtained with MutScape software. 

## Libraries 

```{r, results="hide", message=FALSE, warning=FALSE}
# R version employed for this script is 4.2.0
library(tidyr) # to manage data
library(dplyr) # to manage data
library(RColorBrewer) 
library(rtracklayer)
library(pheatmap) # heatmap
library(clusterProfiler) # enrichment analysis
library(enrichplot)
library(msigdbr) # for enrichment analysis
library(org.Hs.eg.db) # human data base
# library(ggVennDiagram) # venn diagram
library(ggvenn)
library(DOSE)
library(pathview) # visualize pathways
library(simplifyEnrichment)
```

## Upload data

```{r}
# Read results from MutScape significantly mutated genes detection
path <- "/omics/groups/OE0436/data/jimenez/outputs/mutation_analysis_all_patients/MutScape/"
results_resp <- read.delim(paste0(path, "oncodriveclust_results_resp_pre_allmut.tsv"))
results_non <- read.delim(paste0(path, "oncodriveclust_results_nonresp_pre_allmut.tsv"))

# Catch significantly mutated genes in each group
sig_genes_resp <- results_resp %>%
  dplyr::filter(QVALUE < 0.05) %>% dplyr::select(GENE)

sig_genes_non <- results_non %>%
  dplyr::filter(QVALUE < 0.05) %>% dplyr::select(GENE)

# All_sig_genes 
all_sig_genes <- unique(c(sig_genes_resp$GENE, sig_genes_non$GENE))
sig_genes <- data.frame("Gene" = all_sig_genes, 
                        "group" = ifelse(all_sig_genes %in% sig_genes_resp$GENE,
                                         "significant_in_responders",
                                         "significant_in_non_responders"))

# Save genes 
write.table(sig_genes, 
            file = "/omics/groups/OE0436/data/jimenez/outputs/mutation_analysis_all_patients/MutScape/sig_genes_mutscape_pre_respnonresp.tsv", 
            sep = "\t", quote = FALSE, col.names = TRUE)

# Import SNV vcf data and prepare it 
# Read SNV data
snv <- read.delim("/omics/groups/OE0436/data/jimenez/data/snv_indels/vcf/all_snvs_all_patients_metastasis1_definitive.vcf")  

# Read pheno data
pheno <- read.csv(file = "/omics/groups/OE0436/data/jimenez/data/PID_treatment.csv", 
                  header = TRUE, sep = ";", row.names = "PID")
colnames(pheno) = c("therapy", "evolution") # change names

pheno <- pheno %>% 
  mutate(response = ifelse(evolution == "PR", "responders", "non-responders")) %>%
  filter(rownames(pheno) %in%  unique(snv$PID))
```

## Heatmap
Oncoprint makes a better representation of this mutational landscape than the heatmap.

```{r}
# Filter rows that do not represent a somatic mutation and select those interesting fields 
# Also filter for those significantly mutated genes
snv_select <- snv %>% 
  filter(RECLASSIFICATION=="somatic" | RECLASSIFICATION=="somatic_precious") %>%
  dplyr::select(PID, GENE, ANNOVAR_FUNCTION, EXONIC_CLASSIFICATION, X.CHROM, POS)  %>% 
  filter(GENE%in%all_sig_genes, PID%in%rownames(pheno))

# Create a matrix of gene alterations 
matrix <- snv_select %>%
  dplyr::select(PID, GENE) %>%
  dplyr::mutate(value=1) %>%
  dplyr::distinct() %>%
  tidyr::spread(GENE, value)

# Transpose matrix and eliminate PID column
rownames(matrix) <- matrix$PID
matrix <- matrix %>% dplyr::select(-PID) %>% t

# Replace NA by 0
matrix[is.na(matrix)] <- 0

# Subset pheno
pheno <- pheno[colnames(matrix), ]

# Heatmap
pdf("/omics/groups/OE0436/data/jimenez/outputs/mutation_analysis_all_patients/MutScape/heatmaps/heatmap_signmutgenes_pre_all_respnonresp.pdf")

set.seed(123) # for generation of colors 
pheatmap(matrix, scale = 'row',
         annotation_col = pheno[,c(2,3)],
         show_rownames = TRUE, show_colnames = TRUE,
         annotation_legend = TRUE, annotation_names_col = FALSE,
         row_title = "MutScape significantly mutated genes",
         row_title_gp = gpar(fontsize = 10),
         row_names_side = "left",
         column_title = "Metastasis1 samples (pre-treatment)",
         column_title_side = "bottom",
         column_title_gp = gpar(fontsize = 10),
        )

dev.off()
```

## Enrichment analysis

For enrichment analysis, clusterProfiler^1^ package is implemented, which conducts enrichment tests based on hypergeometric distribution and adjusts p-values and computes q-values. It allows for graphical representation of the terms/processes enriched based on p-value, gene counts and GeneRatio (the number of genes in the input list that match with the specific term divided by the number of genes of your gene list annotated in the whole hallmark). 

```{r}
genes <- list("responders" = sig_genes_resp$GENE, "non-responders" = sig_genes_non$GENE)

# Annotate gene symbols as EntrezID
genes_entrez <- lapply(genes, function(x) bitr(x, fromType="SYMBOL", toType="ENTREZID", OrgDb="org.Hs.eg.db"))

# Enrichment 
ego_resp <- enrichGO(gene          = genes_entrez$responders$ENTREZID,
                   keyType       = "ENTREZID",
                   OrgDb         = org.Hs.eg.db,
                   ont           = "BP",
                   pAdjustMethod = "BH",
                   pvalueCutoff  = 0.2,
                   qvalueCutoff  = 0.2,
                   readable      = TRUE)

ek_resp <- enrichKEGG(gene = genes_entrez$responders$ENTREZID,
                    pvalueCutoff = 0.2,
                    qvalueCutoff = 0.2)

ego_non <- enrichGO(gene          = genes_entrez$`non-responders`$ENTREZID,
                   keyType       = "ENTREZID",
                   OrgDb         = org.Hs.eg.db,
                   ont           = "BP",
                   pAdjustMethod = "BH",
                   pvalueCutoff  = 0.4,
                   qvalueCutoff  = 0.5,
                   readable      = TRUE)

ek_non <- enrichKEGG(gene = genes_entrez$`non-responders`$ENTREZID,
                    pvalueCutoff = 0.2,
                    qvalueCutoff = 0.2)


# Table with all of them
ego_resp@result$Response <- "responders"
ego_resp@result$Type <- "GO term" 

ego_non@result$Response <- "non-responders"
ego_non@result$Type <- "GO term" 

ek_resp@result$Response <- "responders"
ek_resp@result$Type <- "KEGG pathway" 

ek_non@result$Response <- "non-responders"
ek_non@result$Type <- "KEGG pathway" 

enrichment <- rbind(ego_resp@result, ego_non@result, ek_resp@result, ek_non@result)

write.table(enrichment, 
            file = "/omics/groups/OE0436/data/jimenez/outputs/mutation_analysis_all_patients/MutScape/enrichment/mutscape_enrichment.csv", quote = FALSE, sep = ",", row.names = TRUE, col.names = TRUE)

# Some information
# For ego_resp
cat("There are", dim(ego_resp@result[ego_resp@result$p.adjust < 0.05, ])[1], 
    "enriched processes in the responders mutated genes with an adjusted p-value < 0.05")

ego_resp@result[ego_resp@result$p.adjust < 0.05, ]
# There don't seem to be other than generic pathways altered

terms <- c("RNA splicing", "mRNA processing")
ego_resp_mRNA <- subset_enrichResult(ego_resp, which(ego_resp@result$Description %in% terms))

# For ek_resp
cat("There are", dim(ek_resp@result[ek_resp@result$p.adjust < 0.05, ])[1], 
    "enriched pathways in the responders mutated genes with an adjusted p-value < 0.05")

ek_resp@result[ek_resp@result$p.adjust < 0.05, ]

#There seem to be a set of altered pathways related to immune functions. They are subset
pathways <- c("Natural killer cell mediated cytotoxicity", 
              "B cell receptor signaling pathway", 
              "T cell receptor signaling pathway",  
              "PD-L1 expression and PD-1 checkpoint pathway in cancer",
              "Chemokine signaling pathway")
ek_resp_immune <- subset_enrichResult(ek_resp, which(ek_resp@result$Description %in% pathways))
ek_resp_immune@result$geneID <- c("NRAS", "NRAS", "NRAS", "NRAS", "NRAS")

# For ego_non
cat("There are", dim(ego_non@result[ego_non@result$p.adjust < 0.15, ])[1], 
    "enriched processes in the non-responders mutated genes with an adjusted p-value < 0.15")

ego_non@result[ego_non@result$p.adjust < 0.15, ]

# There seem to be a set of altered terms related to negative immune functions. They are subset

terms_non <- c("negative regulation of humoral immune response mediated by circulating immunoglobulin",
               "negative regulation of humoral immune response",
               "negative regulation of immunoglobulin mediated immune response",
               "negative regulation of B cell activation",
               "positive regulation of tolerance induction" )
ego_non_immune <- subset_enrichResult(ego_non, which(ego_non@result$Description %in% terms_non))

# For ek_non
cat("There are", dim(ek_non@result[ek_non@result$p.adjust < 0.2, ])[1], 
    "enriched pathways in the non-responders mutated genes with an adjusted p-value < 0.2")

ek_non@result[ek_non@result$p.adjust < 0.2, ]

# Two interesting terms
pathways_non <- c("Leukocyte transendothelial migration", "Natural killer cell mediated cytotoxicity")
ek_non_immune <- subset_enrichResult(ek_non, which(ek_non@result$Description %in% pathways_non))
ek_non_immune@result$geneID <- c("VASP", "BRAF")

# Graphical representations
pdf("/omics/groups/OE0436/data/jimenez/outputs/mutation_analysis_all_patients/MutScape/enrichment/enrichment_analysis_mutscape_sig_all_patients_respnonresp_pre.pdf")

# Responders
enrichplot::dotplot(ego_resp, title = "GO terms encriched in responders significantly mutated genes", showCategory = 10)
cnetplot(ego_resp)
heatplot(ego_resp)
enrichplot::emapplot(pairwise_termsim(ego_resp))
enrichplot::dotplot(ek_resp, title = "KEGG pathways enriched in responders significantly mutated genes", showCategory = 10)

# With selected pathways
enrichplot::dotplot(ek_resp_immune, title = "KEGG immune pathways altered in responders")
cnetplot(ek_resp_immune)
heatplot(ek_resp_immune)
enrichplot::dotplot(ego_resp_mRNA, title = "GO mRNA terms altered in responders")
cnetplot(ego_resp_mRNA)

# Non-responders
enrichplot::dotplot(ego_non, title = "GO terms enriched in non-responders significantly mutated genes", showCategory = 10)
cnetplot(ego_non)
heatplot(ego_non)
emapplot(pairwise_termsim(ego_non))
enrichplot::dotplot(ek_non, title = "KEGG pathways enriched in non-responders significantly mutated genes", showCategory = 10)

# With selected terms and pathways
enrichplot::dotplot(ek_non_immune, title = "KEGG immune pathways altered in non-responders")
cnetplot(ek_non_immune)
heatplot(ek_non_immune)
enrichplot::dotplot(ego_non_immune, title = "GO immune terms altered in non-responders")
cnetplot(ego_non_immune)
heatplot(ego_non_immune)

dev.off()

# KEGG pathway plots

# Responders 

# Natural Killer mediated cytotoxicity
pathview(gene.data = genes_entrez$responders$ENTREZID, pathway.id = "hsa04650", species = "hsa")
pathview(gene.data = genes_entrez$responders$ENTREZID, pathway.id = "hsa04650", 
                species = "hsa", kegg.native = F)

# B cell receptor signalling pathway
pathview(gene.data = genes_entrez$responders$ENTREZID, pathway.id = "hsa04662", species = "hsa")
pathview(gene.data = genes_entrez$responders$ENTREZID, pathway.id = "hsa04662", 
                species = "hsa", kegg.native = F)

# PD-L1 expression and PD-1 checkpoint pathway in cancer
pathview(gene.data = genes_entrez$responders$ENTREZID, pathway.id = "hsa05235", species = "hsa")
pathview(gene.data = genes_entrez$responders$ENTREZID, pathway.id = "hsa05235", 
                species = "hsa", kegg.native = F)

# T cell receptor signaling pathway 
pathview(gene.data = genes_entrez$responders$ENTREZID, pathway.id = "hsa04660", species = "hsa")
pathview(gene.data = genes_entrez$responders$ENTREZID, pathway.id = "hsa04660", 
                species = "hsa", kegg.native = F)


## Venn Diagrams
pdf("/omics/groups/OE0436/data/jimenez/outputs/mutation_analysis_all_patients/MutScape/venn_mutscape_sig_genes.pdf")

ggvenn(genes, fill_color = c("green", "red"))

dev.off()
```

## Significant genes in post-treatment samples 

There were no significant genes detected in responders. In that way, we can say immune checkpoint inhibitors might be targeting those cell lines bearing the previously detected mutations. The detected genes in non-responders were the same.

```{r}
# # Read results from MutScape significantly mutated genes detection
# path <- "/omics/groups/OE0436/data/jimenez/outputs/mutation_analysis_all_patients/MutScape/"
# results_resp_met2 <- read.delim(paste0(path, "oncodriveclust_results_resp_met2_allmut.tsv"))
# results_non_met2 <- read.delim(paste0(path, "oncodriveclust_results_nonresp_pre_allmut.tsv"))
# 
# # Catch significantly mutated genes in each group
# sig_genes_resp_met2 <- results_resp_met2 %>%
#   dplyr::filter(QVALUE < 0.05) %>% dplyr::select(GENE)
# 
# sig_genes_non_met2 <- results_non_met2 %>%
#   dplyr::filter(QVALUE < 0.05) %>% dplyr::select(GENE)
# 
# # All_sig_genes 
# all_sig_genes_met2 <- unique(c(sig_genes_resp_met2$GENE, sig_genes_non_met2$GENE))
# sig_genes_met2 <- data.frame("Gene" = all_sig_genes_met2, 
#                         "group" = ifelse(all_sig_genes_met2 %in% sig_genes_resp_met2$GENE,
#                                          "significant_in_responders",
#                                          "significant_in_non_responders"))
# # sig_genes$group[sig_genes$Gene == "BRAF"] <- "significant_in_both"
# 
# # Save genes 
# write.table(sig_genes_met2, 
#             file = "/omics/groups/OE0436/data/jimenez/outputs/mutation_analysis_all_patients/MutScape/sig_genes_mutscape_met2_respnonresp.tsv", 
#             sep = "\t", quote = FALSE, col.names = TRUE)
# 
# ## Venn Diagrams
# genes <- list("responders metastasis1" = sig_genes_resp$GENE, 
#               "responders metastasis2" = sig_genes_resp_met2$GENE,
#               "non-responders metastasis1" = sig_genes_non$GENE, 
#               "non-responders metastasis2" = sig_genes_non_met2$GENE)
# 
# pdf("/omics/groups/OE0436/data/jimenez/outputs/mutation_analysis_all_patients/MutScape/venn_mutscape_sig_genes_prepost.pdf")
# 
# ggvenn(genes, fill_color = c("green", "red", "light green", "orange"), 
#        text_size = 4, set_name_size = 2) 
# 
# 
# dev.off()
```

## References
1. 	Yu, G., Wang, L.-G., Han, Y. & He, Q.-Y. clus-terProfiler: an R Package for Comparing Bio-logical Themes Among Gene Clusters. doi:10.1089/omi.2011.0118 (2012).